Non-word repetition in Yélî Dnye

### 2020-12-22 manuscript.tex
update to-do notes
cleaned up and reknitted

\setmainfont{Doulos SIL} % Set main font to Doulos SIL

\title{Non-word repetition in Yélî Dnye}
\author{Alejandrina Cristia\textsuperscript{1}~\& Marisa
Casillas\textsuperscript{2,3}}
\date{}

\authornote{All data are made available in a repository in the
Open Science Framework. AC acknowledges the support of the Agence
Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004,
ANR-14-CE30-0003, ANR-17-EURE-0017); and the J. S. McDonnell Foundation
Understanding Human Cognition Scholar Award.

Correspondence concerning this article should be addressed to
Alejandrina Cristia, 29, rue d'Ulm, 75005 Paris, France. E-mail:
\href{mailto:alecristia@gmail.com}{\nolinkurl{alecristia@gmail.com}}}

\abstract{
In nonword repetition (NWR) studies, participants are presented
auditorily with an item that is phonologically legal but lexically
meaningless in their language. NWR scores are thought to reflect
long-term phonological knowledge as well as online phonological working
memory and flexible production patterns. In this study, we report on NWR
results among children learning Yélî Dnye, an isolate spoken on Rossel
Island in Papua New Guinea that has an unusually dense phonological
inventory. This study contributes to three lines of research. First, we
document that non-word items containing typologically frequent sounds
are repeated without changes more often that non-words containing
typologically rare sounds. Second, we document rather weak effects of
item length, contributing to other research suggesting that length
effects may be language-specific. Third, we do not find strong
individual variation effects in this population, contrary to previous
results documenting strong age-related effects. Together, these data
provide a unique view of online phonological processing in an
understudied language while making preliminary connections between
patterns in language development and in the frequency of linguistic
features in human language at large.

@@

\subsection{TODO Alex}\label{todo-alex}

\begin{itemize}
\tightlist
\item
  check again if all fields are explained
\item
  set up OSF (or revise if already there)
\item
  implement decisions on prop phones \& patterns nwr mp
\item
  revise bullet points for discussion
\end{itemize}

\subsection{TODO Middy}\label{todo-middy}

\begin{itemize}
\tightlist
\item
  draft discussion
\end{itemize}

\subsection{Introduction}\label{introduction}

@@

NWR has been used to seek answers to a variety of theoretical questions,
including what the links are between phonology, working memory, and the
lexicon (Bowey, 2001), and how extensively phonological constraints
found in the lexicon affect online production (Gallagher, 2014). NWR is
also frequently used in applied contexts, notably as a diagnostic tool
for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
Since non-words can be generated in any language, it has attracted the
attention of researchers working in multilingual and linguistically
diverse environments, particularly in Europe (Action, 2009; Meir,
Walters, \& Armon-Lotem, 2016). However, it has seldom been used outside
of Europe and North America (with exceptions including Gallagher, 2014,
and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)). Further,
the languages it has been used with have historically had small to
moderate phonological inventories (Maddieson, 2005).

@@

\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}

Yélî Dnye is an isolate language (presumed Papuan) spoken by
approximately 7,000 people residing on Rossel Island, an island found at
the far end of the Louisiade Archipelago in Milne Bay Province, Papua
New Guinea. The Yélî sound system, much like its baroque grammatical
system (Levinson, accepted), is unlike any other in the region.

With only four primary places of articulation (bilabial, alveolar,
post-alveolar, and velar) and no voicing contrasts, the phonological
inventory is remarkably packed with acoustically similar segments. The
core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
substantial portion of them contrastively pre-nasalized or nasally
released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/,
/ṭpṇm/, /kpŋm/). A large number of this combinatorial set can further be
contrastively labialized, palatalized on release, or both (e.g., /pʲ/,
/pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (accepted) for
details).\footnote{We use Levinson's (accepted) under-dot notation
  (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these
  stops are, articulatorily, somewhat variable in place, with at least
  some tokens produced fully sub-apically. In approximating
  cross-linguistic segment frequency below we use the corresponding
  retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The
consonantal inventory also includes a number of non-nasal continuants
(/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye may be oral
or nasal, short or long. The 10 oral vowel qualities, which span four
levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/,
/ɑ/) can be produced as short and long vowels, with 7 of these able to
appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/,
/æ̃/, /ɑ̃/).

In total, Yélî Dnye uses 90 distinctive segments (not including an
additional 3 rarely used consonants), far outstripping the phonemic
inventory size of other documented Papuan languages (Foley, 1986;
Levinson, accepted) and includes at least two contrasts not yet been
documented elsewhere (labial-coronal double-articulations with dental
vs.~post-alveolar coronal placement for both oral and nasal stops,
Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, in preparation).

Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
that, and with tri-and-above syllabic words appearing least frequently
(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
recent Yélî Dnye dictionary at time of writing; Levinson, accepted). The
vast majority of syllables use a CV format. A small portion of the
lexicon features words with a final closed syllable, limited to codas of
-/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
coda, yielding a CV.CV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
process is used frequently with English loan words that have a coda
(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
these are limited to /æ/-. Finally, the lexicon features a handful of
single-vowel grammatical morphemes (see Levinson (accepted) for
details).

\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}

Most speakers of Yélî Dnye grow up speaking it monolingually until they
begin attending school around the age of 7; school instruction is in
English. While monolingual Yélî Dnye upbringing is common, multilingual
families are not unusual, particularly in the region around the Catholic
Mission---the same region in which the current data were
collected---where there is a higher incidence of married-in mothers from
other islands (Brown \& Casillas, in press). Children in these
multilingual families grow up speaking Yélî Dnye plus English, Tok
Pisin, and/or another language(s) from the region.

@@

Despite restricted outside contact, formal education is a priority for
many Yélî families. A recent study surveying more than 40 parents of
young children in this region of the island found that nearly all
parents completed 6 or more years of education, with 65\% of mothers and
49\% of fathers completing the full range of locally available education
(\textasciitilde{}9 years), with around half of those parents having
pursued an additional 2+ years of education on other islands in the
region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
there are a handful of local schools around the island, these may be
well out of walking distance for many children (i.e., more than 1 hour
on foot or by canoe each day). It is very common for households that are
situated close to a school to billet their school-aged relatives during
the weekdays (sometimes the weekends too, if the child's home is very
far) for long segements of each school year. Combined with practices of
collective child guardianship within each close-knit hamlet, the
practical consequence of this billeting for researchers is that adult
consent can often come from a combination of aunts, uncles, adult
cousins, and grandparents standing in for the child's biological
parents. On top of this, child assent is culturally pertinent, as child
independence is encouraged and respected from toddlerhood onward (Brown
\& Casillas, in press).\footnote{For these reasons, we were unable to
  record precise details for a few children, including child date of
  birth, parent age in years, and parent education.} Finally, as
mentioned, the language of schooling in this region is English, and so
the majority of teens and adults on the island speak some English---many
speak it very fluently.

While work on Yélî language development is growing (e.g., Brown, 2011,
2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
linguistic development is quite limited, and research into their
phonological development in particular has only just begun (e.g., Peute,
Fikkert, \& Casillas, In preparation). What we do know about the early
language environment is that, while direct speech to children under 3;0
is relatively infrequent throughout the day (Casillas et al., 2020),
shared caregiving practices and a near-universal fondness for social
interaction with young children ensure that speech directed to children
comes from all types of speakers: women, men, and other children (Bunce
et al., under review; Casillas et al., 2020). While speech from women
predominates in young children's speech environments on Rossel, as it
does elsewhere (Bergelson et al., 2019; Bunce et al., under review),
there is a significant and marked increase in child-directed speech from
other children as infants get older (Bunce et al., under review;
Casillas et al., 2020). This increase in child-directed speech from
other children is attributed to the fact that, starting around age 2,
children often spend large swaths of the day playing, swimming, and
foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10
members) independent and mixed-age child play groups (Brown \& Casillas,
in press; Casillas et al., 2020).

\subsubsection{NWR design and analysis
adaptations}\label{nwr-design-and-analysis-adaptations}

@@

A second list of candidate items was generated in 2019 by selecting
complex consonants and systematically crossing them with all the vowels
in the Yélî Dnye inventory to produce consonant-vowel monosyllabic
forms. As before, items were automatically excluded if they appeared in
the dictionary. Additionally, since perceiving vowel length in isolated
monosyllables is challenging, any item that had a short/long lexical
neighbor was excluded. Because there is still much to discover about the
phonology and phonetics of Yélî Dnye (Levinson, accepted), it was also
possible that we initially generated items with illegal, but currently
undocumented constraints. Therefore, we made sure that the precise
consonant-vowel sequence occurred in some real word in the dictionary
(i.e., that there was a longer word included the monosyllable as a
subsequence). These candidates were then presented to one informant, for
a final check that they did not mean anything. Together with the 2018
selection, they were recorded, based on their orthographic forms, using
a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832
stereo audio recorder (using an XLR to mini-jack adapter) by the same
informant, monitored by the second author for clear production of the
phonological target. The complete recorded list was finally presented to
two more informants, who were able to repeat all the items and who
confirmed there were no real words present. Despite these checks, one
monosyllable was ultimately frequently identified as a real word in the
resulting data (intended ``yî'' /yɯ/; identified as ``yi'' /yi/,
\emph{tree}). This item is removed from the analyses below.

@@

We ran children in batches, testing within a handful of hamlet clusters
spread across the northeastern region of the island. Because space
availability was limited in different ways from hamlet to hamlet, the
places where elicitation happened varied across testing sites. We tested
in four different sites, only making a single visit to each, conducting
back-to-back testing of all eligible children present at the time of our
visit in order to prevent the items from `spreading' between children
through hearsay. In the first hamlet, we tested children in five
different places, with some children being tested inside a house and
others tested on the veranda. The complete list of places and the ways
in which they met the desiderata mentioned above can be found in the raw
data, available from the online supplementary materials. Whenever
children living in the same household were tested, we tried to test
children in age order, from oldest to youngest, to minimize intimidation
for younger household members, and always using different elicitation
sets.

@@

Additionally, we estimated the typological frequency of all phonological
segments used in the target items using the PHOIBLE cross-linguistic
phonological inventory database (Moran \& McCloy, 2019). For each phone
in our task, we extracted the number and percentage of languages noted
to have that phone in its inventory. While PHOIBLE is an unprecedentedly
comprehensive database, with phonological inventory data for over 2000
languages at time of writing, it is of course still far from complete,
so estimated frequencies should be taken with a pinch of salt. In both
PHOIBLE and UPSID---another large phonemic inventory database---note
that nearly half of the segment types are only attested in one language
(Steven Moran, personal communication; see also
\href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}).
It is therefore unsurprising that we did not find three of our segments
(/lβʲ/, /ṭp/, and /tp/) in the database. Following the general pattern
that many phones only appear once cross-linguistically, even in large
databases, we treat our ``missing'' phones as having a frequency of 1,
with a (rounded) percentage of 0\%.

@@

We tested a total of 55 children from 38 families spread across four
hamlets. We excluded test sessions from analysis for the following
reasons: refused participation or failure to repeat items presented over
headphones even after coaching (N=8), spoke too softly to allow offline
coding (N=5), or were 13 years old or older (N=2; we tested these
teenagers to put younger children at ease). The remaining 40 children
(14 girls) were aged 6.98 years (range 3.92-11.03 years). Among these
children were 34 only exposed to Yélî Dnye at home and 6 children
exposed to Yélî Dnye plus one or more other languages at home. Maternal
years of education averaged 8.22 years (range 6-12 years).\footnote{Years
  of education is difficult to precisely pinpoint in this context.
  Schooling beyond grade 8 takes place off-island and enrollment is tied
  to an ability to pay that year's school fees, seasonal interruptions
  to school opening times, familial obligations, and more. We asked
  parents to report their highest completed level of education. We then
  record the number of years entailed by having completed that level
  under ideal conditions. This method results in many esimates around
  typical ``graduation'' times (e.g., graduating from primary education
  before starting secondary education).} In terms of birth order, 6 were
first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth.

\subsection{Results}\label{results}

\subsubsection{Preliminary analyses}\label{preliminary-analyses}

@@

\subsubsection{NWR as a function of cross-linguistic phone
frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}

@@

We could include 850 observations, from 40 children producing in any
given trial one of 41 potential target words. The analysis revealed a
main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant
estimate for the scaled average cross-linguistic frequency of phones in
the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target
words with phones found more frequently across languages had a higher
proportion of words that were correctly repeated, as shown in Figure
\ref{fig:Fig2-xling-freq}. Averaging across items and participants, the
Pearson correlation between scaled average cross-linguistic phone
frequency and whole-item NWR scores was r(31) = 0.63.

@@

\subsubsection{NWR scores as a function of item
length}\label{nwr-scores-as-a-function-of-item-length}

@@

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}
\begin{tabular}{lll}
\toprule
Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
\midrule
47 (22) & 59 (17) & 41 (17)\\
79 (22) & 92 (9) & 8 (9)\\
78 (19) & 93 (7) & 7 (7)\\
74 (32) & 91 (12) & 9 (12)\\
\bottomrule
\end{tabular}
\end{threeparttable}
\end{center}
\end{table}

\subsubsection{Factor structuring individual
variation}\label{factor-structuring-individual-variation}

@@

\subsubsection{Item complexity}\label{item-complexity}

>>>>

\setmainfont{Doulos SIL} % Set main font to Doulos SIL
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\makeatother
\shorttitle{NWE in Yélî Dnye}
\author{Alejandrina Cristia\textsuperscript{1}\ \& Marisa Casillas\textsuperscript{2,3}}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} Laboratoire de Sciences Cognitives et de Psycholinguistique, Département d'Etudes cognitives, ENS, EHESS, CNRS, PSL University\\\textsuperscript{2} Max Planck Institute for Psycholinguistics\\\textsuperscript{3} University of Chicago}
\authornote{All data are made available in a repository in the Open Science Framework. AC acknowledges the support of the Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004, ANR-14-CE30-0003, ANR-17-EURE-0017); and the J. S. McDonnell Foundation Understanding Human Cognition Scholar Award.

Correspondence concerning this article should be addressed to Alejandrina Cristia, 29, rue d’Ulm, 75005 Paris, France. E-mail: alecristia@gmail.com}
\keywords{phonology, non-word repetition, development\newline\indent Word count: ~xxx words}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=english]{babel}
\else
  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\fi

\title{Non-word repetition in Yélî Dnye}

\date{}

\abstract{
In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language. NWR scores are thought to reflect long-term phonological knowledge as well as online phonological working memory and flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea that has an unusually dense phonological inventory. This study contributes to three lines of research. First, we document that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds. Second, we document rather weak effects of item length, contributing to other research suggesting that length effects may be language-specific. Third, we do not find strong individual variation effects in this population, contrary to previous results documenting strong age-related effects. Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between patterns in language development and in the frequency of linguistic features in human language at large.
}

@@

\hypertarget{todo-alex}{%
\subsection{TODO Alex}\label{todo-alex}}

\begin{itemize}
\tightlist
\item
  check whether it's stimuli, stimuli2, stimuli3 that should be used!!!
\item
  create fields final-data DONE
\item
  leave only files actually needed DONE
\item
  add readme explaining files DONE
\item
  remove all files that aren't necessary DONE
\item
  get rid of all the conversions and directly read in xls
\item
  set up OSF (or revise if already there)
\item
  implement decisions on prop phones \& patterns nwr mp
\item
  check and accept/deny MC's updated segment frequencies
\item
  re-run analyses w/ new frequencies and check that outcomes didn't qualitatively change
\item
  check YD data in figure 1 (the effect of item length looks reversed)
\item
  see MC comments throughout and delete now-irrelevant things
\end{itemize}

\hypertarget{todo-middy}{%
\subsection{TODO Middy}\label{todo-middy}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

@@

NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe (Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has seldom been used outside of Europe and North America (with exceptions including Gallagher, 2014, and @cristia2020infant). Further, the languages it has been used with have historically had small to moderate phonological inventories (Maddieson, 2005).

@@

\hypertarget{yuxe9luxee-dnye-phonology}{%
\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}}

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system (Levinson, n.d.), is unlike any other in the region.

With only four primary places of articulation (bilabial, alveolar, post-alveolar, and velar) and no voicing contrasts, the phonological inventory is remarkably packed with acoustically similar segments. The core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a substantial portion of them contrastively pre-nasalized or nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can further be contrastively labialized, palatalized on release, or both (e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (n.d.) for details).\footnote{We use Levinson's (n.d.) under-dot notation (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these stops are, articulatorily, somewhat variable in place, with at least some tokens produced fully sub-apically. In approximating cross-linguistic segment frequency below we use the corresponding retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The consonantal inventory also includes a number of non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short and long vowels, with 7 of these able to appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/, /æ̃/, /ɑ̃/).

In total, Yélî Dnye uses 90 distinctive segments (not including an additional 3 rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages (Foley, 1986; Levinson, n.d.) and includes at least two contrasts not yet been documented elsewhere (labial-coronal double-articulations with dental vs.~post-alveolar coronal placement for both oral and nasal stops, Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, n.d.).

Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with monosyllabic words (\textasciitilde{}40\%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently (\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; Levinson, n.d.). The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., \enquote{ndap} (Spondylus shell) /ṇṭæp/). However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., \enquote{ndapî} /\enquote{ṇṭæ.pɯ/). This process is used frequently with English loan words that have a coda (e.g., 'ponî} (phone) /\enquote{pɔ.nɯ/). An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala} (here) /æ.'læ/), but these are limited to /æ/-. Finally, the lexicon features a handful of single-vowel grammatical morphemes (see Levinson (n.d.) for details).

\hypertarget{the-yuxe9luxee-community}{%
\subsubsection{The Yélî community}\label{the-yuxe9luxee-community}}

Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands (Brown \& Casillas, n.d.). Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or another language(s) from the region.

@@

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65\% of mothers and 49\% of fathers completing the full range of locally available education (\textasciitilde{}9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward (Brown \& Casillas, n.d.).\footnote{For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.} Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

While work on Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, n.d.; Casillas et al., 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's linguistic development is quite limited, and research into their phonological development in particular has only just begun (e.g., Peute, Fikkert, \& Casillas, n.d.). What we do know about the early language environment is that, while direct speech to children under 3;0 is relatively infrequent throughout the day (Casillas et al., 2020), shared caregiving practices and a near-universal fondness for social interaction with young children ensure that speech directed to children comes from all types of speakers: women, men, and other children (Bunce et al., n.d.; Casillas et al., 2020). While speech from women predominates in young children's speech environments on Rossel, as it does elsewhere (Bergelson et al., 2019; Bunce et al., n.d.), there is a significant and marked increase in child-directed speech from other children as infants get older (Bunce et al., n.d.; Casillas et al., 2020). This increase in child-directed speech from other children is attributed to the fact that, starting around age 2, children often spend large swaths of the day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10 members) independent and mixed-age child play groups (Brown \& Casillas, n.d.; Casillas et al., 2020).

\hypertarget{nwr-design-and-analysis-adaptations}{%
\subsubsection{NWR design and analysis adaptations}\label{nwr-design-and-analysis-adaptations}}

@@

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yélî Dnye inventory to produce consonant-vowel monosyllabic forms. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since perceiving vowel length in isolated monosyllables is challenging, any item that had a short/long lexical neighbor was excluded. Because there is still much to discover about the phonology and phonetics of Yélî Dnye (Levinson, n.d.), it was also possible that we initially generated items with illegal, but currently undocumented constraints. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were then presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded, based on their orthographic forms, using a Shure SM10A XLR dynamic headband microphone and an Olympus WS-832 stereo audio recorder (using an XLR to mini-jack adapter) by the same informant, monitored by the second author for clear production of the phonological target. The complete recorded list was finally presented to two more informants, who were able to repeat all the items and who confirmed there were no real words present. Despite these checks, one monosyllable was ultimately frequently identified as a real word in the resulting data (intended \enquote{yî} /yɯ/; identified as \enquote{yi} /yi/, \emph{tree}). This item is removed from the analyses below.

@@

We ran children in batches, testing within a handful of hamlet clusters spread across the northeastern region of the island. Because space availability was limited in different ways from hamlet to hamlet, the places where elicitation happened varied across testing sites. We tested in four different sites, only making a single visit to each, conducting back-to-back testing of all eligible children present at the time of our visit in order to prevent the items from \enquote{spreading} between children through hearsay. In the first hamlet, we tested children in five different places, with some children being tested inside a house and others tested on the veranda. The complete list of places and the ways in which they met the desiderata mentioned above can be found in the raw data, available from the online supplementary materials. Whenever children living in the same household were tested, we tried to test children in age order, from oldest to youngest, to minimize intimidation for younger household members, and always using different elicitation sets.

@@

Additionally, we estimated the typological frequency of all phonological segments used in the target items using the PHOIBLE cross-linguistic phonological inventory database (Moran \& McCloy, 2019). For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory. While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at time of writing, it is of course still far from complete, so estimated frequencies should be taken with a pinch of salt. In both PHOIBLE and UPSID---another large phonemic inventory database---note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication; see also \href{http://web.phonetik.uni-frankfurt.de/upsid_info.html}{web.phonetik.uni-frankfurt.de/upsid\_info.html}). It is therefore unsurprising that we did not find three of our segments (/lβʲ/, /ṭp/, and /tp/) in the database. Following the general pattern that many phones only appear once cross-linguistically, even in large databases, we treat our \enquote{missing} phones as having a frequency of 1, with a (rounded) percentage of 0\%.

@@

We tested a total of 55 children from 38 families spread across four hamlets. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=8), spoke too softly to allow offline coding (N=5), or were 13 years old or older (N=2; we tested these teenagers to put younger children at ease). The remaining 40 children (14 girls) were aged 6.98 years (range 3.92-11.03 years). Among these children were 34 only exposed to Yélî Dnye at home and 6 children exposed to Yélî Dnye plus one or more other languages at home. Maternal years of education averaged 8.22 years (range 6-12 years).\footnote{Years of education is difficult to precisely pinpoint in this context. Schooling beyond grade 8 takes place off-island and enrollment is tied to an ability to pay that year's school fees, seasonal interruptions to school opening times, familial obligations, and more. We asked parents to report their highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions. This method results in many esimates around typical \enquote{graduation} times (e.g., graduating from primary education before starting secondary education).} In terms of birth order, 6 were first borns, 5 second, 2 third, 7 forth, 5 fifth, and 1 sixth.

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{preliminary-analyses}{%
\subsubsection{Preliminary analyses}\label{preliminary-analyses}}

@@

\hypertarget{nwr-as-a-function-of-cross-linguistic-phone-frequency}{%
\subsubsection{NWR as a function of cross-linguistic phone frequency}\label{nwr-as-a-function-of-cross-linguistic-phone-frequency}}

@@

We could include 850 observations, from 40 children producing in any given trial one of 41 potential target words. The analysis revealed a main effect of age (ß = 0.33, SE ß = 0.13, p = 0.01); and a significant estimate for the scaled average cross-linguistic frequency of phones in the target words (ß = 0.82, SE ß = 0.18, p \textless{} 0.001): Target words with phones found more frequently across languages had a higher proportion of words that were correctly repeated, as shown in Figure \ref{fig:Fig2-xling-freq}. Averaging across items and participants, the Pearson correlation between scaled average cross-linguistic phone frequency and whole-item NWR scores was r(32) = 0.61.

@@

\hypertarget{nwr-scores-as-a-function-of-item-length}{%
\subsubsection{NWR scores as a function of item length}\label{nwr-scores-as-a-function-of-item-length}}

@@

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:tablength}NWR measured in whole-word scores, phoneme-based scores, and normalized Levenshtein Distance, separately for the four stimuli lengths.}

\begin{tabular}{lll}
\toprule
Word & \multicolumn{1}{c}{Phoneme} & \multicolumn{1}{c}{NLD}\\
\midrule
47 (22) & 59 (17) & 41 (17)\\
79 (22) & 92 (9) & 8 (9)\\
78 (19) & 93 (7) & 7 (7)\\
74 (32) & 91 (12) & 9 (12)\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

\hypertarget{factor-structuring-individual-variation}{%
\subsubsection{Factor structuring individual variation}\label{factor-structuring-individual-variation}}

@@

\hypertarget{item-complexity}{%
\subsubsection{Item complexity}\label{item-complexity}}
